---
title: "Traumatic Brain Injury"
author: "Zenthia Song"
output: github_document
---

## Get the data

```{r}
# Load the necessary libraries
library(tidyverse)
library(dplyr)
library(tidyr)
library(gridExtra)
library(ggtext)

# Loading data
tbi_age <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_age.csv", show_col_types = FALSE)
tbi_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_year.csv', show_col_types = FALSE)
tbi_military <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_military.csv', show_col_types = FALSE)
```

## Explore the data

```{r}
# Summary and check the data
glimpse(tbi_age)
glimpse(tbi_year)
glimpse(tbi_military)
```

```{r}
# Explore number estimates by age group and injury mechanism
tbi_age_num <- tbi_age %>%
  group_by(age_group, injury_mechanism) %>%
  summarise(total_number_est = sum(number_est, na.rm = TRUE), .groups = "drop")

tbi_age_num

# Explore number and rate estimates by year and injury mechanism
tbi_year_num <- tbi_year %>%
  group_by(year, injury_mechanism) %>%
  summarise(total_number_est = sum(number_est, na.rm = TRUE), .groups = "drop")

tbi_year_num

# Explore number of diagnoses by year, service, and severity
tbi_military_num <- tbi_military %>%
  group_by(year, service, severity) %>%
  summarise(total_diagnosed = sum(diagnosed, na.rm = TRUE), .groups = "drop")

tbi_military_num

```

## RQ 1: What is the distribution of TBI severity (mild, moderate, severe, penetrating) in different branches of the military? 

### Tidying data

```{r}
# Select the necessary columns and filter for 'Active' component
tbi_military_active <- tbi_military %>%
  filter(severity != "Not Classifiable" & component == "Active") %>%
  select(service, severity, diagnosed)

# For the 'Active' component, group by 'service' and 'severity', and calculate the total number of diagnosed cases
tbi_active_grouped <- tbi_military_active %>%
  group_by(service, severity) %>%
  summarise(total_diagnosed = sum(diagnosed, na.rm = TRUE), .groups = "drop")

# Do the same for the 'Guard' component
tbi_military_guard <- tbi_military %>%
  filter(severity != "Not Classifiable" & component == "Guard") %>%
  select(service, severity, diagnosed)

# Do the same for the 'Guard' component
tbi_guard_grouped <- tbi_military_guard %>%
  group_by(service, severity) %>%
  summarise(total_diagnosed = sum(diagnosed, na.rm = TRUE), .groups = "drop")

# Do the same for the 'Reserve' component
tbi_military_Reserve <- tbi_military %>%
  filter(severity != "Not Classifiable" & component == "Reserve") %>%
  select(service, severity, diagnosed)

# Do the same for the 'Reserve' component
tbi_Reserve_grouped <- tbi_military_Reserve %>%
  group_by(service, severity) %>%
  summarise(total_diagnosed = sum(diagnosed, na.rm = TRUE), .groups = "drop")
```

### Visualizing Data

```{r}
# Bar plot of total_diagnosed by service and severity for 'Active' component
p1 <- ggplot(tbi_active_grouped, aes(x = service, y = total_diagnosed, fill = severity)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Service", y = "Total Diagnosed", fill = "Severity",
       title = "Number of Diagnosed TBI Cases by Severity and Service in Active Component")

# Do the same for the 'Guard' component
p2 <- ggplot(tbi_guard_grouped, aes(x = service, y = total_diagnosed, fill = severity)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Service", y = "Total Diagnosed", fill = "Severity",
       title = "Number of Diagnosed TBI Cases by Severity and Service in Guard Component")

# Do the same for the 'Reserve' component
p3 <- ggplot(tbi_Reserve_grouped, aes(x = service, y = total_diagnosed, fill = severity)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Service", y = "Total Diagnosed", fill = "Severity",
       title = "Number of Diagnosed TBI Cases by Severity and Service in Reserve Component")

# Arrange the plots in a grid
combined_plot <- grid.arrange(p1, p2, p3, ncol = 2, heights = c(3, 3, 3))

# Save the plot as a PNG file
ggsave("traumatic-brain-injury_1.png", combined_plot, width = 15, height = 13)
```

## RQ 2: What is the annual trend in the number of deaths due to different injury mechanisms causing traumatic brain injury? 

### Tidying data

```{r}
# Filter the data for 'Deaths' type and "Total group"
tbi_year_deaths <- tbi_year %>%
  filter(injury_mechanism != "Total" & type == "Deaths") %>%
  select(year, injury_mechanism, number_est)

# Group the data by 'year' and 'injury_mechanism', and calculate the total number of cases
tbi_year_deaths_grouped <- tbi_year_deaths %>%
  group_by(year, injury_mechanism) %>%
  summarise(total_deaths = sum(number_est, na.rm = TRUE), .groups = "drop")
```

### Visualizing Data

```{r}
# Line plot of total_deaths by year and injury_mechanism
ggplot(tbi_year_deaths_grouped, aes(x = year, y = total_deaths, color = injury_mechanism)) +
  geom_line() +
  labs(x = "Year", y = "Total Deaths", color = "Injury Mechanism",
       title = "Total Number of Deaths by Injury Mechanism Over Time")

# Save the plot as a PNG file
ggsave("traumatic-brain-injury_2.png", width = 8, height = 6)
```

## RQ 3: What are the primary causes of traumatic brain injury (TBI)? And how do they affect different age groups?

### Tidying data

```{r}
# Define injury mechanism recode
injury_mechanism_recodes <- list( "Unintentional Falls" = "Unintentional Fall", "Intentional self-harm" = "Intentional Self-harm", "Motor Vehicle Crashes" = "Motor Vehicle Crash", "Unintentionally struck by or against an object" = "Unintentionally Struck by an Object", "Other unintentional injury, mechanism unspecified" = "Other/Unknown", "Other or no mechanism specified" = "Other/Unknown")

# Filtering and recoding data
tbi_age <- tbi_age %>%
  filter(age_group != "Total", age_group != "0-17", !is.na(number_est)) %>%
  mutate(type = recode(type, "Emergency Department Visit" = "Non-lethal TBI (98% of cases)", "Hospitalizations" = "Non-lethal TBI (98% of cases)", "Deaths" = "Lethal TBI (2% of cases)")) %>%
  group_by(type, injury_mechanism, age_group) %>%
  summarize(number_est = sum(number_est), .groups = "drop") %>%
  ungroup() %>%
  mutate(
    injury_mechanism = recode(injury_mechanism, !!! injury_mechanism_recodes),
    injury_mechanism = fct_relevel(injury_mechanism, rev(c("Intentional Self-harm", "Assault", "Motor Vehicle Crash", "Unintentional Fall", "Unintentionally Struck by an Object", "Other/Unknown"))),
    age_group = fct_relevel(age_group, c("0-4", "5-14", "15-24", "25-34", "35-44", "45-54", "55-64", "65-74", "75+"))
  )

tbi_age
```

### Visualizing Data

```{r}
# Create a dataframe with labels for each injury mechanism and the proportion it represents
injury_labels <- tribble(
  ~type, ~injury_mechanism, ~age_group, ~prop,
  "Lethal TBI (2% of cases)", "Intentional Self-harm", "25-34", 0.2,
  "Lethal TBI (2% of cases)", "Assault", "25-34", 0.5,
  "Lethal TBI (2% of cases)", "Motor Vehicle Crash", "5-14", 0.66,
  "Lethal TBI (2% of cases)", "Unintentional Fall", "45-54", 0.8,
  "Lethal TBI (2% of cases)", "Other/Unknown", "45-54", 0.94,
  "Non-lethal TBI (98% of cases)", "Assault", "25-34", 0.07,
  "Non-lethal TBI (98% of cases)", "Motor Vehicle Crash", "15-24", 0.26,
  "Non-lethal TBI (98% of cases)", "Unintentional Fall", "35-44", 0.55,
  "Non-lethal TBI (98% of cases)", "Unintentionally Struck by an Object", "0-4", 0.79,
  "Non-lethal TBI (98% of cases)", "Other/Unknown", "15-24", 0.93
)

# Define colors for the plot
background_color <- "white"
text_color <- "grey15"
selfharm_color <- "#FFD700"
assault_color <- "#FFB14E"
mvcrash_color <- "#FA8775"
unfall_color <- "#EA5F94"
unstruck_color <- "#CD34B5"
other_color <- "#9D02D7"

# Create the plot
ggplot() +
  geom_col(data = tbi_age, aes(x = age_group, y = number_est, fill = injury_mechanism), position = "fill", width = 1) +
  geom_text(data = injury_labels, aes(x = age_group, y = prop, label = injury_mechanism), color = "white", size = 4.5, family = "Times New Roman", hjust = 0) +
  facet_wrap(. ~ type, ncol = 2) +
  scale_x_discrete(labels = c(" 5", seq(15, 75, by = 10), "")) +
  scale_fill_manual(values = rev(c(selfharm_color, assault_color, mvcrash_color, unfall_color, unstruck_color, other_color))) +
  guides(fill = "none") +
  labs(title = "In 2014, the United States recorded 2.87 million cases of traumatic brain injury (TBI)",
       x = "Age group", y = NULL) +
  
   # Set the theme for the plot
  theme_minimal(base_family = "Times New Roman", base_size = 18) +
  theme(
    # Define the appearance of the plot title
    plot.title = element_text(family = "Times New Roman", size = 24, color = text_color),
    plot.title.position = "plot",
    # Define the appearance of the plot background
    plot.background = element_rect(fill = background_color, color = background_color),
    plot.margin = margin(1.5, 1.5, 1, 1.5, unit = "line"),
    # Remove the gridlines
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank(),
    # Define the appearance of the text in the plot
    strip.text = element_text(color = text_color, size = 16),
    axis.title = element_text(color = text_color),
    axis.text.x = element_text(color = text_color, hjust = -1.6, vjust = 5),
    axis.text.y = element_blank()
  )

# Save the plot as a PNG file
ggsave("traumatic-brain-injury_3.png", width = 12, height = 6)
```

